---
title: "Final Project Computational Analysis and Figures"
output: word_document
date: "2025-04-07"
---

Import Cleaned data sets
```{r}
# Utilizing personal paths for cleaned files
flu_data <- read.csv(file = "cleaned_flu_dataset.csv")

vax_data <-read.csv(file = "cleaned_vax_dataset_FINAL.csv")

```


Flu Data sorted by State (region in .csv file) and by Flu season (season_description in .csv file)

```{r}
# Sort flu data by state (region in .csv file), and flu season (season_description in .csv file))
sorted_flu_data <- flu_data[order(flu_data$region, flu_data$season_description), ]

# Total cases of flu data
  # Adds up all confirmed flu sub types 
flu_data$total_cases <- with(flu_data,
  a_2009_h1n1 + a_h3 + a_subtyping_not_performed +
  b + b_vic + b_yam + h3n2v + a_h5
)

# Confirms total cases column worked properly
  # Output should display first couple of values in new column
head(flu_data$total_cases)

# Total flu cases by state and season
state_season_totals <- aggregate(
  cbind(total_cases, total_specimens) ~ region + season_description,
  data = flu_data,
  FUN = sum,
  na.rm = TRUE # missing values won't mess up totals
)

print(head(state_season_totals))

print(head(sorted_flu_data))

```
```{r}
summary(state_season_totals)
```

```{r}
library(writexl)

write_xlsx(state_season_totals, "flu_state_and_season_totals.xlsx")
```


Vax Data Sorted by State and Season (2024-2025)

```{r}
# Sort vax_data by state (jurisdiction) and flu season (current_season)
sorted_vax_data <- vax_data[order(vax_data$Jurisdiction, vax_data$Current_season), ]

# Total vax cases by state and season
vax_totals <- aggregate(
  cbind(Numerator, Population) ~ Jurisdiction + Current_season,
  data = sorted_vax_data,
  FUN = sum,
  na.rm = TRUE # missing values won't mess up totals
)

colnames(vax_totals) <- c("state", "season", "total_vaccinated", "total_population")

print(head(vax_totals))
```

```{r}
summary(vax_totals)
```


Vax Data Sorted by State and Season (2022-2023 & 2023-2024)
```{r}
sorted_prev_vax_data <- vax_data[order(vax_data$Jurisdiction, vax_data$Previous_flu_Season), ]

# Get total vaccinated for each state and previous season
prev_vax_totals <- aggregate(
  Previous_season_numerator ~ Jurisdiction + Previous_flu_Season,
  data = sorted_prev_vax_data,
  FUN = sum,
  na.rm = TRUE
)

# Rename columns for clarity
colnames(prev_vax_totals) <- c("state", "season", "total_vaccinated_prev")


print(head(prev_vax_totals))
```

```{r}
summary(prev_vax_totals)
```


Vax Summary Stats - Current Season (2024-2025)

```{r}
# Summary statistics for current season vaccinations
current_vax_summary <- aggregate(
  Numerator ~ Jurisdiction + Current_season,
  data = vax_data,
  FUN = function(x) c(
    total = sum(x, na.rm = TRUE),
    average = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE)
  )
)

# Flatten the summary table
current_vax_summary <- data.frame(
  Jurisdiction = current_vax_summary$Jurisdiction,
  Current_season = current_vax_summary$Current_season,
  total_vax = current_vax_summary$Numerator[, "total"],
  avg_vax = current_vax_summary$Numerator[, "average"],
  median_vax = current_vax_summary$Numerator[, "median"],
  min_vax = current_vax_summary$Numerator[, "min"],
  max_vax = current_vax_summary$Numerator[, "max"],
  sd_vax = current_vax_summary$Numerator[, "sd"]
)

print(head(current_vax_summary)) # print first few lines
```

Send Current Season Vax Summary Stats to an Excel File for easy viewing purposes
```{r}
library(writexl)

write_xlsx(current_vax_summary, "current_season_vax_summary.xlsx")
```


Vax Summary Stats - Previous Season (2022-2023 & 2023-2024)
```{r}
# Summary statistics for previous season vaccinations
prev_vax_summary <- aggregate(
  Previous_season_numerator ~ Jurisdiction + Previous_flu_Season,
  data = vax_data,
  FUN = function(x) c(
    total = sum(x, na.rm = TRUE),
    average = mean(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    min = min(x, na.rm = TRUE),
    max = max(x, na.rm = TRUE),
    sd = sd(x, na.rm = TRUE)
  )
)

# Flatten the summary table
prev_vax_summary <- data.frame(
  Jurisdiction = prev_vax_summary$Jurisdiction,
  Previous_flu_Season = prev_vax_summary$Previous_flu_Season,
  total_vax_prev = prev_vax_summary$Previous_season_numerator[, "total"],
  avg_vax_prev = prev_vax_summary$Previous_season_numerator[, "average"],
  median_vax_prev = prev_vax_summary$Previous_season_numerator[, "median"],
  min_vax_prev = prev_vax_summary$Previous_season_numerator[, "min"],
  max_vax_prev = prev_vax_summary$Previous_season_numerator[, "max"],
  sd_vax_prev = prev_vax_summary$Previous_season_numerator[, "sd"]
)

print(head(prev_vax_summary)) # print first few lines

```

Send Previous Season Vax Summary Stats to an Excel File for easy viewing purposes
```{r}
library(writexl)

write_xlsx(prev_vax_summary, "previous_season_vax_summary.xlsx")
```


Merge the Vax and Flu totals for further analysis

```{r}
# --- Clean flu data ---

# Update flu data - column names with similar information need to match
flu_data_clean <- flu_data[flu_data$region_type == "States", ]  # keep only states
flu_data_clean$season <- gsub("Season ", "", flu_data_clean$season_description)  # change seaso format to "202x-2x"
flu_data_clean$state <- flu_data_clean$region  # rename region to state

# Aggregate flu totals by state and season
flu_totals <- aggregate(
  cbind(total_specimens, a_2009_h1n1, a_h3, a_subtyping_not_performed, b, b_vic, b_yam, h3n2v, a_h5) ~ state + season,
  data = flu_data_clean,
  FUN = sum,
  na.rm = TRUE
)

# Add the total flu cases column (sum of all flu strains)
flu_totals$total_cases <- rowSums(flu_totals[, c("a_2009_h1n1", "a_h3", "a_subtyping_not_performed", "b", "b_vic", "b_yam", "h3n2v", "a_h5")], na.rm = TRUE)

# --- Clean vaccination data ---

# Extract 2022–23 and 2023–24 data from previous seasons
prev_vax_data_clean <- sorted_prev_vax_data
prev_vax_data_clean$season <- prev_vax_data_clean$Previous_flu_Season
prev_vax_data_clean$state <- prev_vax_data_clean$Jurisdiction

# Extract 2024–25 data from current season
curr_vax_data_clean <- sorted_vax_data
curr_vax_data_clean$season <- curr_vax_data_clean$Current_season
curr_vax_data_clean$state <- curr_vax_data_clean$Jurisdiction

# Combine both
vax_data_clean <- rbind(
  prev_vax_data_clean[, c("state", "season", "Numerator", "Population")],
  curr_vax_data_clean[, c("state", "season", "Numerator", "Population")]
)

# Aggregate vaccinations
vax_totals <- aggregate(
  cbind(Numerator, Population) ~ state + season,
  data = vax_data_clean,
  FUN = sum,
  na.rm = TRUE
)

# Calculate vaccination rate
vax_totals$total_vaccinated <- vax_totals$Numerator
vax_totals$total_population <- vax_totals$Population
vax_totals$vaccination_rate <- vax_totals$total_vaccinated / vax_totals$total_population

# --- Merge flu and vaccination data ---
merged_data <- merge(
  vax_totals[, c("state", "season", "total_vaccinated", "total_population", "vaccination_rate")],
  flu_totals[, c("state", "season", "total_cases", "total_specimens")],
  by = c("state", "season")
)

# Preview
head(merged_data)
```

Send Merged Data to an Excel File for easy viewing purposes
```{r}
library(writexl)

write_xlsx(merged_data, "merged_data.xlsx")
```

Correlation Analysis and Linear Regression 

```{r}
# Checking correlation
cor(merged_data$vaccination_rate, merged_data$total_cases, use = "complete.obs")
```
```{r}
# Value greater than 0, less than 1, suggests weak positive correlation
# run cor.test to determine p value of this correlation
cor.test(merged_data$vaccination_rate, merged_data$total_cases)
```
```{r}
# Use ggplot() to visualize correlation
library(ggplot2)
ggplot(merged_data, aes(x=vaccination_rate, y=total_cases)) + geom_point() + geom_smooth(method="lm") + labs(title="Vaccination Rate vs Total Flu Cases")
```
```{r}
# Setup linear regression model
linearModel <- lm(total_cases ~ vaccination_rate, data = merged_data)
summary(linearModel)
```
```{r}
# Plot linearModel
plot(linearModel)
```
```{r}
# Generate a second linear model, this time incorporate total_population and total_specimens, to control for those factors.
populationControlled_linearModel <- lm(total_cases ~ vaccination_rate + total_population + total_specimens, data = merged_data)
# print summary
summary(populationControlled_linearModel)
```
```{r}
# p value is more significant.
# plot this population controlled model.
plot(populationControlled_linearModel)
```


Separate High vs. Low vaccination rates
```{r}
# Use median to define High vs. Low
threshold <- median(merged_data$vaccination_rate, na.rm = TRUE)

merged_data$vaccination_group <- ifelse(
  merged_data$vaccination_rate >= threshold,
  "High",
  "Low"
)

```


Hypothesis Testing - Welch's T-test

```{r}
seasons <- unique(merged_data$season)

# Initialize list to store results
t_test_results <- list()

# Loop through each season and perform t-test
for (szn in seasons) {
  season_data <- merged_data[merged_data$season == szn, ]
  test_result <- t.test(total_cases ~ vaccination_group, data = season_data)
  t_test_results[[szn]] <- test_result
  
  cat("\nSeason:", szn, "\n")
  print(test_result)
}
```


Scatter Plot of Total Flu Cases vs Vaccination Rate
```{r}
# Scatter Plot

# 2022-23
plot_2022_23 <- ggplot(subset(merged_data, season == "2022-23"), aes(x = vaccination_rate, y = total_cases, color = state)) +
  geom_point(size = 3) +
  labs(
    title = "Total Flu Cases vs. Vaccination Rate (2022-23)",
    x = "Vaccination Rate",
    y = "Total Flu Cases",
    color = "State"
  ) +
  theme_minimal()

# 2023-24
plot_2023_24 <- ggplot(subset(merged_data, season == "2023-24"), aes(x = vaccination_rate, y = total_cases, color = state)) +
  geom_point(size = 3) +
  labs(
    title = "Total Flu Cases vs. Vaccination Rate (2023-24)",
    x = "Vaccination Rate",
    y = "Total Flu Cases",
    color = "State"
  ) +
  theme_minimal()

# 2024-25
plot_2024_25 <- ggplot(subset(merged_data, season == "2024-25"), aes(x = vaccination_rate, y = total_cases, color = state)) +
  geom_point(size = 3) +
  labs(
    title = "Total Flu Cases vs. Vaccination Rate (2024-25)",
    x = "Vaccination Rate",
    y = "Total Flu Cases",
    color = "State"
  ) +
  theme_minimal()

plot_2022_23
plot_2023_24
plot_2024_25
```

Scatter Plots with regression lines for High v Low vaccination groups
```{r}
# Scatter plot with regression lines for High vs. Low vaccination groups
library(ggplot2)

ggplot(merged_data, aes(x = vaccination_rate, y = total_cases, color = vaccination_group)) +
  geom_point(size = 3) +  # Scatter points
  geom_smooth(method = "lm", aes(group = vaccination_group), se = FALSE) +  # Regression lines
  labs(
    title = "Total Flu Cases vs. Vaccination Rate (High vs. Low Vaccination Groups)",
    x = "Vaccination Rate",
    y = "Total Flu Cases",
    color = "Vaccination Group"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

```{r}
high_model <- lm(total_cases ~ vaccination_rate, data = subset(merged_data, vaccination_group == "High"))
low_model <- lm(total_cases ~ vaccination_rate, data = subset(merged_data, vaccination_group == "Low"))

# Extract and print slopes (coefficients for vaccination_rate)
high_slope <- coef(high_model)[["vaccination_rate"]]
low_slope <- coef(low_model)[["vaccination_rate"]]

cat("Slope for High vaccination group:", round(high_slope, 2), "\n")
cat("Slope for Low vaccination group:", round(low_slope, 2), "\n")
```

```{r}
# Scatter plot with regression lines for High vs. Low vaccination groups
# Create plots for each season
seasons <- unique(merged_data$season)

for (s in seasons) {
  season_data <- subset(merged_data, season == s)
  
  p <- ggplot(season_data, aes(x = vaccination_rate, y = total_cases, color = vaccination_group)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", aes(group = vaccination_group), se = FALSE) +
    labs(
      title = paste("Flu Cases vs. Vaccination Rate - Season:", s),
      x = "Vaccination Rate",
      y = "Total Flu Cases",
      color = "Vaccination Group"
    ) +
    theme_minimal() +
    theme(legend.position = "top")
  
  print(p)
}
```

```{r}

for (s in seasons) {
  season_data <- subset(merged_data, season == s)
  
  cat("\nSeason:", s, "\n")
  
  for (group in c("High", "Low")) {
    group_data <- subset(season_data, vaccination_group == group)
    model <- lm(total_cases ~ vaccination_rate, data = group_data)
    slope <- coef(model)[["vaccination_rate"]]
    cat("  Slope for", group, "vaccination group:", round(slope, 2), "\n")
  }
}
```



